---
title: "Random_forest_workshop"
output: html_document
date: "2024-03-07"
refrence: "https://www.kaggle.com/code/mrisdal/exploring-survival-on-the-titanic/report"
---
first we clear all variables from the current R session.
```{r}
rm(list=ls())
```

Load Libraries
if the libraries are not installed you can use the command install.packages("package_name") to install it.
```{r}
library('ggplot2') # visualization
library('ggthemes') # visualization
library('scales') # visualization
library('dplyr') # data manipulation
library('mice') # imputation
library('randomForest') # classification algorithm
```

Load and check input data
```{r}
train <- read.csv('./train.csv', stringsAsFactors = F)
test  <- read.csv('./test.csv', stringsAsFactors = F)

full  <- bind_rows(train, test) # bind training & test data
head(full)
```


Feature Engineering

```{r}
# Grab title from passenger names
full$Title <- gsub('(.*, )|(\\..*)', '', full$Name)

# Show title counts by sex
table(full$Sex, full$Title)

```

```{r}
# Titles with very low cell counts to be combined to "rare" level
rare_title <- c('Dona', 'Lady', 'the Countess','Capt', 'Col', 'Don', 
                'Dr', 'Major', 'Rev', 'Sir', 'Jonkheer')

# Also reassign mlle, ms, and mme accordingly
full$Title[full$Title == 'Mlle']        <- 'Miss' 
full$Title[full$Title == 'Ms']          <- 'Miss'
full$Title[full$Title == 'Mme']         <- 'Mrs' 
full$Title[full$Title %in% rare_title]  <- 'Rare Title'

# Show title counts by sex again
table(full$Sex, full$Title)
```
```{r}
# Finally, grab surname from passenger name
full$Surname <- sapply(full$Name,  
                      function(x) strsplit(x, split = '[,.]')[[1]][1])
```

```{r}
# Create a family size variable including the passenger themselves
full$Fsize <- full$SibSp + full$Parch + 1

# Create a family variable 
full$Family <- paste(full$Surname, full$Fsize, sep='_')
head(full)
```

Do families sink or swim together?

Now that we've extracted meaningful information from the 'Passenger Name' variable, we can proceed to create additional family-related variables. We will start by generating a 'Family Size' variable based on the number of siblings/spouse and the number of children/parents.

```{r}
# Use ggplot2 to visualize the relationship between family size & survival
ggplot(full[1:891,], aes(x = Fsize, fill = factor(Survived))) +
  geom_bar(stat='count', position='dodge') +
  scale_x_continuous(breaks=c(1:11)) +
  labs(x = 'Family Size') +
  theme_few()
```
Observing a survival penalty for both single passengers and those with family sizes above 4, we can simplify this variable into three levels. This simplification will be beneficial since there are relatively fewer large families. Let's proceed to create a discretized family size variable

```{r}
# Discretize family size
full$FsizeD[full$Fsize == 1] <- 'singleton'
full$FsizeD[full$Fsize < 5 & full$Fsize > 1] <- 'small'
full$FsizeD[full$Fsize > 4] <- 'large'
```

Here’s probably some potentially useful information in the passenger cabin variable including about their deck. Let’s take a look.

```{r}
# This variable appears to have a lot of missing values
full$Cabin[1:28]
```


```{r}
# The first character is the deck. For example:
strsplit(full$Cabin[2], NULL)[[1]]
```
```{r}
# Create a Deck variable. Get passenger deck A - F:
full$Deck<-factor(sapply(full$Cabin, function(x) strsplit(x, NULL)[[1]][1]))
full$Deck[1:28]
```

Missing Values:

Now that we've completed the initial data preparation steps, we can move on to exploring and addressing missing data through imputation. With the dataset being relatively small, deleting entire observations (rows) or variables (columns) with missing values may not be the best approach. Instead, we're left with the option of replacing these missing values with sensible estimates based on the data's distribution, such as using the mean, median, or mode. Additionally, we can consider prediction methods. We'll apply both of these techniques, and I'll rely on data visualization to help guide our decisions.

```{r}
# Passengers 62 and 830 are missing Embarkment
full[c(62, 830), 'Embarked']
full[62, ]
full[830, ]
```
We will infer their values for embarkment based on present data that we can imagine may be relevant: fare and passenger class


```{r}
# Get rid of our missing passenger IDs
embark_fare <- full %>%
  filter(PassengerId != 62 & PassengerId != 830)

# Use ggplot2 to visualize embarkment, passenger class, & median fare
ggplot(embark_fare, aes(x = Embarked, y = Fare, fill = factor(Pclass))) +
  geom_boxplot() +
  geom_hline(aes(yintercept=80), 
    colour='red', linetype='dashed', lwd=2) +
  scale_y_continuous(labels=dollar_format()) +
  theme_few()
```

The median fare paid by a first-class passenger departing from Cherbourg ('C') aligns well with the $80 fare paid by our passengers with missing embarkment information. It seems reasonable to replace the NA values with 'C'.

```{r}
# Since their fare was $80 for 1st class, they most likely embarked from 'C'
full$Embarked[c(62, 830)] <- 'C'
```

We’re close to fixing the handful of NA values here and there. Passenger on row 1044 has an NA Fare value.

```{r}
# Show row 1044
full[1044, ]
```

This is a third class passenger who departed from Southampton (‘S’). Let’s visualize Fares among all others sharing their class and embarkment (n = 494).

```{r}
ggplot(full[full$Pclass == '3' & full$Embarked == 'S', ], 
  aes(x = Fare)) +
  geom_density(fill = '#99d6ff', alpha=0.4) + 
  geom_vline(aes(xintercept=median(Fare, na.rm=T)),
    colour='red', linetype='dashed', lwd=1) +
  scale_x_continuous(labels=dollar_format()) +
  theme_few()
```
From this visualization, it seems quite reasonable to replace the NA Fare value with median for their class and embarkment which is $8.05.

```{r}
# Replace missing fare value with median fare for class/embarkment
full$Fare[1044] <- median(full[full$Pclass == '3' & full$Embarked == 'S', ]$Fare, na.rm = TRUE)
```

T here are quite a few missing Age values in our data.We will create a model predicting ages based on other variables.


```{r}
# Show number of missing Age values
sum(is.na(full$Age))
```

We can use mice package to predict missing ages.The mice package in R stands for "Multivariate Imputation by Chained Equations." It is a package used for multiple imputation of missing data. Multiple imputation is a technique used when data is missing in a dataset. The mice package allows you to impute missing values in a dataset by creating multiple imputations using chained equations.

```{r}
# Make variables factors into factors
factor_vars <- c('PassengerId','Pclass','Sex','Embarked',
                 'Title','Surname','Family','FsizeD')

full[factor_vars] <- lapply(full[factor_vars], function(x) as.factor(x))

# Set a random seed
set.seed(129)

# Perform mice imputation, excluding certain less-than-useful variables:
mice_mod <- mice(full[, !names(full) %in% c('PassengerId','Name','Ticket','Cabin','Family','Surname','Survived')], method='rf') 
```

```{r}
# Save the complete output 
mice_output <- complete(mice_mod)
```

```{r}
# Plot age distributions
par(mfrow=c(1,2))
hist(full$Age, freq=F, main='Age: Original Data', 
  col='darkgreen', ylim=c(0,0.04))
hist(mice_output$Age, freq=F, main='Age: MICE Output', 
  col='lightgreen', ylim=c(0,0.04))
```
Now we replace our age vector in the original data with the output from the mice model.
```{r}
# Replace Age variable from the mice model.
full$Age <- mice_output$Age

# Show new number of missing Age values
sum(is.na(full$Age))
```
Now that we know everyone’s age, we can create a couple of new age-dependent variables: Child and Mother. A child will simply be someone under 18 years of age and a mother is a passenger who is female, is over 18, has more than 0 children, and does not have the title ‘Miss’

```{r}
# First we'll look at the relationship between age & survival
ggplot(full[1:891,], aes(Age, fill = factor(Survived))) + 
  geom_histogram() + 
  # I include Sex since we know (a priori) it's a significant predictor
  facet_grid(.~Sex) + 
  theme_few()
```


```{r}
# Create the column child, and indicate whether child or adult
full$Child[full$Age < 18] <- 'Child'
full$Child[full$Age >= 18] <- 'Adult'

# Show counts
table(full$Child, full$Survived)
```

Prediction

```{r}
# Split the data back into a train set and a test set
train <- full[1:891,]
test <- full[892:1309,]
```

```{r}
full$Child  <- factor(full$Child)
```


```{r}
# Set a random seed
set.seed(754)

# Build the model (note: not all possible variables are used)
rf_model <- randomForest(factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + 
                                            Fare + Embarked + Title + 
                                            FsizeD + Child ,
                                            data = train)

# Show model error
plot(rf_model, ylim=c(0,0.36))
legend('topright', colnames(rf_model$err.rate), col=1:3, fill=1:3)
```
The black line shows the overall error rate which falls below 20%. The red and green lines show the error rate for ‘died’ and ‘survived’ respectively.

Variable importance:
Let’s look at relative variable importance by plotting the mean decrease in Gini calculated across all trees.

```{r}
# Get importance
importance    <- importance(rf_model)
varImportance <- data.frame(Variables = row.names(importance), 
                            Importance = round(importance[ ,'MeanDecreaseGini'],2))

# Create a rank variable based on importance
rankImportance <- varImportance %>%
  mutate(Rank = paste0('#',dense_rank(desc(Importance))))

# Use ggplot2 to visualize the relative importance of variables
ggplot(rankImportance, aes(x = reorder(Variables, Importance), 
    y = Importance, fill = Importance)) +
  geom_bar(stat='identity') + 
  geom_text(aes(x = Variables, y = 0.5, label = Rank),
    hjust=0, vjust=0.55, size = 4, colour = 'red') +
  labs(x = 'Variables') +
  coord_flip() + 
  theme_few()
```
we can see from the graph that our title variable has highest relative importance out of all of our predictor variables.\

Prediction:

```{r}
# Predict using the test set
prediction <- predict(rf_model, test)

# Save the solution to a dataframe with two columns: PassengerId and Survived (prediction)
solution <- data.frame(PassengerID = test$PassengerId, Survived = prediction)

# Write the solution to file
write.csv(solution, file = 'rf_mod_Solution.csv', row.names = F)
```













